Geometric models overestimate lake depth due to imperfect slope measurement

### 2020-07-28 manuscript/agujournaltemplate.tex
use pjh for initials
fill out acknowledgements, closes #13
prep for submission

\begin{document}

%% ------------------------------------------------------------------------ %%
%  Title
%
% (A title should be specific, informative, and brief. Use
% abbreviations only if they are defined in the abstract. Titles that
% start with general keywords then specific terms are optimized in
% searches)
%
%% ------------------------------------------------------------------------ %%

@@

\begin{keypoints}
\item Geometric models to predict lake depth, which require in-lake slope, assume that nearshore land slope is a good proxy for in-lake slope.
\item Using data from thousands of lakes, we show that nearshore land slope is a poor proxy for in-lake slope and increases prediction error.
\item Increases in prediction error were systematic such that depth was overpredicted in concave and reservoir lakes.
\end{keypoints}

@@

% The Plain Language Summary should be written for a broad audience,
% including journalists and the science-interested public, that will not have
% a background in your field.
%
% A Plain Language Summary is required in GRL, JGR: Planets, JGR: Biogeosciences,
% JGR: Oceans, G-Cubed, Reviews of Geophysics, and JAMES.
% see http://sharingscience.agu.org/creating-plain-language-summary/)
%
%% ------------------------------------------------------------------------ %%

%% \begin{abstract} starts the second page

\begin{abstract}
      Lake depth is a critical characteristic that influences many important
      ecological processes in lakes. Unfortunately, lake depth measurements
      are labor-intensive to gather and are only available for a small
      fraction of lakes globally. Therefore, scientists have tried to predict
      lake depth from characteristics that are easily obtained for all lakes
      such as lake surface area or the slope of the land surrounding a lake.
      One approach for predicting lake depth simulates lake basins using a
      geometric model where nearshore land slope is assumed to be a
      representative proxy for in-lake slope and the distance to the center of
      the lake is assumed to be a representative proxy for the distance to the
      deepest point of the lake. However, these assumptions have rarely been
      tested in a broad range of lakes. We used bathymetry data from 5,000
      lakes and reservoirs to test these assumptions and to examine whether
      differences in lake type or shape influences depth prediction error. We
      found that nearshore land slope was not a representative proxy of
      in-lake slope and using it for prediction increases prediction error
      substantially relative to models using true in-lake slope. We also found
      that models using nearshore land slope as a proxy systematically
      overpredicts lake depth in concave lakes (i.e. bowl-shaped; up to 18\%
      of lakes in the study population) and reservoir lakes (up to 30\% of
      lakes in the study population), suggesting caution in using geometric
      models for depth prediction in unsampled lakes.
\end{abstract}

@@

\section{Introduction}
\noindent
\noindent
Lake depth is an important factor controlling lake physics, chemistry, and biota. Deeper lakes generally have higher water clarity and less complete mixing compared to shallow lakes \cite{feeEffectsLakeSize1996, readSimulating2368Temperate2014}. These differences are reflected in variation among lakes in terms of biological productivity \cite{qinWaterDepthUnderpins2020} and rates of greenhouse gas production \cite{liSignificantContributionLake2020}. However, because measured depth data is only available for a small fraction ($\sim$25\%) of all lakes, our ability to understand and predict depth-dependent processes is limited. The importance of lake depth, coupled with its limited availability, has led to numerous attempts to predict depth using measures available for all lakes such as lake surface area or the nearshore slope of the land surrounding a lake \cite{heathcotePredictingBathymetricFeatures2015, oliver2016prediction, sobekPredictingDepthVolume2011}. Given the limited prediction accuracy of prior attempts ($\pm$ 6-7 m), studies have explored strategies such as employing more diverse covariates \cite{oliver2016prediction}, varying lake buffer sizes \cite{heathcotePredictingBathymetricFeatures2015}, or estimating hidden groupings (e.g. fitting different models for distinct size classes) among lakes \cite{sobekPredictingDepthVolume2011}.

One approach for predicting lake depth involves using a geometric model that assumes lake basins correspond to an idealized shape such as a cone, bowl, or an elliptic sinusoid \cite{hollisterPredictingMaximumLake2011, neumannMaximumDepthAverage1959, yigzawNewGlobalStorage2018}. All such geometric models for lake depth prediction involve implicit assumptions about the terms of geometric formulae. In the simplest case, where lakes basins are treated as cones (Eq \ref{eq1}, Figure \ref{fig1}), two assumptions are required to make depth predictions for all lakes: 1) that nearshore land slope is a representative proxy for in-lake slope and 2) that the distance to the center of the lake is a representative proxy for the distance to the deepest point of the lake (Figure \ref{fig1}). This cone model imposes the following fixed (i.e. geometric) relationship between slope and horizontal distance:

@@

The assumptions of the cone model (as well as other geometric models) can be tested by comparing proxy measures of lake geometry against corresponding “true” (i.e. in-lake) values derived from bathymetric maps and by evaluating how lake cross-section shapes differ from that of an idealized cone \cite{johanssonNewApproachesModelling2007}. For instance, lake cross-section shapes have been shown to vary from narrow "convex" forms to outstretched "concave" forms \cite{hakansonLakeFormLake1977}. Because tests of geometric model assumptions require bathymetric map data, which is only available for about 15\% of all lakes, existing evidence may not be applicable to all lakes. The few studies that have tested these assumptions have mostly been limited to individual studies of very large ($>$ 500 ha) lakes or studies on small numbers ($<$ 100) of lakes \cite{johanssonNewApproachesModelling2007}. Studies focused specifically on reservoirs (as opposed to the more typical case where reservoirs and natural lakes are combined), have been even more restricted to that of extremely large lakes $>$ 1000 ha \cite{lehnerHighresolutionMappingWorld2011, messagerEstimatingVolumeAge2016}.

As a result of this limited testing, we lack knowledge on both the predictive performance of geometric models, the effect of proxies on depth prediction, and whether depth predictions are more sensitive to measurement errors in the horizontal dimension (i.e. distance to the deepest point of the lake) or measurement errors in the vertical dimension (i.e. in-lake slope). Additionally it is unclear whether model prediction error is related to differences in lake type such those with different cross-section shapes or those classified as reservoirs versus natural lakes.

Given the knowledge gaps identified above, we asked three research questions: 1) How representative is nearshore land slope of in-lake slope; and how representative is the distance to the center of a lake compared to the distance to the deepest point of a lake? 2) How does the use of proxies for lake geometry affect lake depth prediction error? 3) How does lake cross-section shape (i.e. concave versus convex) and lake type (i.e. natural lake vs reservoir) affect depth prediction error? To answer these questions, we extracted maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, cross-section shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 lakes in the Northeastern and Midwestern US from bathymetric map data and supplemented this data with classification estimates of whether lakes are reservoirs or natural lakes. First, we examined whether measures of lake geometry (in-lake slope and distance to the deepest point of lakes) were related to geometry proxies (nearshore land slope and distance to lake centers). Next, we computed geometric depth estimates (Equation \ref{eq1}) and prediction "offsets" to these estimates using the random forest algorithm (Equation \ref{eq2}). Covariates used in offset modeling included a variety of lake, watershed, and hydrologic subbasin measures that are available for all lakes (Table \ref{table1}). We examined differences in overall prediction error corresponding to different inputs to geometric depth as well as differences in the relative distribution of prediction error in lakes with different characteristics.

Given that, by definition, the distance proxy (distance to the center of the lake) must always be greater or equal to the true distance value (distance to the deepest point of the lake), we expect that the use of this proxy will lead to overestimation of lake depth (Figure \ref{fig3}). Furthermore we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \cite{lanza1985interactions}. In a similar fashion, we expect to see overestimation error associated with using a nearshore land slope proxy in lakes with differing cross-section shape (Figure \ref{fig2}) such that the depth of U-shaped (i.e. concave) lakes will be overpredicted whereas the depth of V-shaped (i.e. convex) lakes will be underpredicted. Finally, we expect that overall depth prediction will be strongly related to lake area and hydrologic subbasin variables as these measures have been influential in prior studies \cite{oliver2016prediction}.

By testing these expectations, we can establish whether barriers to increased depth prediction accuracy lie in lack of correspondence between true and proxy measures of lake geometry or in hidden grouping (such as lake cross-section shape or reservoir status). This information could help direct future research efforts to focus on particular dimensions of lake geometry (i.e. horizontal versus vertical) or to stratify model predictions based on specific lake types and cross-section shapes. Ultimately, achieving increased depth prediction accuracy would allow for more precise estimates of depth-dependent biotic and chemical processes across broad spatial extents.

\begin{figure}
      \begin{center}
        \includegraphics[width=0.75\textwidth,keepaspectratio]{../figures/lake_shape}
      \end{center}
      \caption{Diagram showing our expectation that slope-based models of lake depth will under predict true depth in convex lakes (left) and over predict true depth in concave lakes (right). Dashed lines represent extrapolated nearshore land slope while solid lines represent the lake bottom.}\label{fig2}
    \end{figure}

\section{Methods}
\subsection{Data description}
\noindent
We compiled bathymetry data on approximately 5,000 lakes in the Northeastern and Midwestern US from nine official state databases (Figure S1). The original data came in a variety of formats including pre-interpolated rasters (Minnesota), contour lines (Nebraska, Michigan, Massachusetts, Kansas, Iowa), contour polygons (New Hampshire, Connecticut), or point depth soundings (Maine). For the Minnesota data, we simply clipped the raster for each lake to its outline. For data from the remaining states, we processed each lake by converting its original representation to a point layer (if necessary), rasterizing these points, and creating an interpolated bathymetry “surface” using a simple moving window average in the \texttt{raster} R package \cite{hijmansRasterGeographicData2019}. The size of the moving window was adjusted iteratively to ensure that each bathymetry raster contained no missing data.

@@

We categorized lakes based on their cross-section shape and reservoir class. For cross-section shape, we categorized lakes as either convex or concave following the method of \citeA{hakansonLakeFormLake1977} by computing normalized lake depth-area relationships (i.e. hypsographic curves) and assigning class membership based on whether a lake’s curve falls above or below that of a simple straight-sided cone (Figure S4). We further classified lakes using data from \citeA{polus2020}, which uses the output of a machine learning algorithm to assign a probability to each lake as to whether it is a reservoir or a natural lake. For our purposes, we determined a lake to be a reservoir if the classification probability was 0.75 or greater. The \citeA{polus2020} data product defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water, beyond simply controlling water level. It makes no distinction between different dam types or dam heights.

\begin{table}
\caption{Summary of lake characteristics for the present study (and for lakes in the contiguous United States from \citeA{stachelek2020}). Predictor variables for computing random forest offsets (Eq \ref{eq2}) are printed in bold face. Dashes (-) indicate an identical sample size among this study and that of the contiguous United States.} \label{table1}
\centering
\setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{lllll}
  \hline
  Variable & Median & Q25 & Q75 & n\\
  \hline
  Max depth (m) & 8.2 (7) & 4.6 (3.7) & 14 (12) & 4850 (17700)\\
  \textbf{Elevation (m)} & 300 (340) & 180 (210) & 400 (460) & 4850 (17700)\\
  \textbf{Area (ha)} & 55 (33) & 21 (11) & 140 (100) & 4850 (17700)\\
  \textbf{Island area (ha)} & 0 (0) & 0 (0) & 0.18 (0.076) & 4850 (17700)\\
  \textbf{Perimeter (m)} & 4400 (3500) & 2500 (1800) & 8100 (7300) & 4850 (17700)\\
  \textbf{Shoreline development} & 1.7 (1.7) & 1.4 (1.4) & 2.1 (2.2) & 4850 (17700)\\
  \textbf{Watershed-lake ratio} & 7.8 (10) & 3.9 (4.4) & 17 (28) & 4850 (17700)\\
  Distance to deepest point (m) & 180 (-) & 110 (-) & 290 (-) & 4850 (-)\\
  Distance to lake center (m) & 240 (-) & 160 (-) & 380 (-) & 4850 (-)\\
  In-lake slope (m/m) & 0.046 (-) & 0.024 (-) & 0.079 (-) & 4850 (-)\\
  Nearshore land slope (m/m) & 0.077 (-) & 0.051 (-) & 0.11 (-) & 4850 (-)\\
  \hline
  \end{tabular}
\end{table}

Covariates used in random forest modeling (Table \ref{table1}, Equation \ref{eq2}) for lake elevation, area, island area, perimeter, shoreline development, watershed to lake area ratio, and hydrologic subbasin (i.e. HUC4s), were obtained from the LAGOS-US LOCUS database \cite{smith2020}. One such measure, that of shoreline development, is a measure of lake perimeter shape defined as:

@@

\noindent
where sinuous lakes have larger values of shoreline development and circular lakes have smaller values of shoreline development. Watershed to lake area ratio is an approximation of water residence time and is defined as watershed area divided by lake area. All of our data and data processing code are available at [Zenodo DOI].

\subsubsection{Proxy evaluation}
\noindent
We conducted a qualitative assessment of whether or not proxy measures of lake geometry are representative of their true values by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination ($R^2$). We further tested proxy measures by examining their effect on lake depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of lake depth using only geometry information ($depth_{geometric}$, Equation \ref{eq1}). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all lakes (Table \ref{table1}). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of $depth_{geometric}$ and among different lake types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table \ref{table2}).

@@

\subsubsection{Random forest models}
\noindent
Nearly all prior studies predicting lake depth using geometric models include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy. For our purposes, this offset modeling enabled us to test our expectations that prediction error would be different among different formulations of $depth_{geometric}$ and among different lake types. It also facilitated direct comparison against prior models of lake depth including those that are non-geometric. We generated an “offset” to geometric depth \cite<sensu>{hollisterPredictingMaximumLake2011} using the random forest algorithm and the \texttt{ranger} R package \cite{wrightRangerFastImplementation2017} to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation \ref{eq1}) along with the lake elevation, area, perimeter, and ratio/index measures listed in Table 1:

@@

\subsubsection{Model evaluations}
\noindent
We evaluated model fit and prediction error using root-mean-square error (RMSE) and coefficient of determination ($R^2$) metrics on a holdout set containing 25\% of all lakes. We evaluated the residuals (residual = observed - predicted) of each model relative to lake cross-section shape and reservoir classes to determine whether depth is consistently over or under predicted for some lake types relative to others. All of our code for model fitting and evaluation is available at [Zenodo DOI].

\section{Results}
\noindent
Lakes belonging to each cross-section shape and reservoir class were not evenly distributed across our study area (Figure S1). For example, concave lakes were nearly absent from Michigan whereas Maine lakes had an overabundance of lakes categorized as neither concave nor convex. Lakes in the southern portions of our study area tended to be classified as reservoirs whereas lakes in the northern portions of our study area were a more even mix between reservoirs and natural lakes (Figure S1). Approximately 18\%, 80\%, and 2\% of lakes were classified as having a concave, convex, or neither shape respectively whereas approximately 30\% and 70\% of lakes were classified as being a reservoir or a natural lake.

Although proxy distance to lake center was often much larger in magnitude compared to the true distance to the deepest point of lakes’, they were strongly related ($R^2$ = 0.8). In contrast, proxy nearshore land slope and true in-lake slope were more weakly related ($R^2$ = 0.17). For slope measures, most lakes had lower magnitude (i.e. shallower) nearshore land slope compared to true in-lake slope (Figure \ref{fig3}). Taken together, these results suggest that proxy distance to the center of lakes is representative of true distance to the deepest point of lakes while proxy nearshore land slope is not representative of true in-lake slope. In addition to overall differences between slope and distance measures, we found differences in these relationships among lake shape classes. For example, in-lake slope and distance to the deepest point of the lake metrics were consistently larger in magnitude for convex lakes as compared to concave lakes (Figure S5). However,  there were not similar differences among slope and distance metrics for natural lakes versus reservoirs (Figure S5).

@@

\begin{table}
  \caption{Model fit and predictive accuracy metrics (RMSE = root mean square error, $R^2$ = coefficient of determination) for all combinations of true (in-lake slope, distance to the deepest point of the lake) and proxy (nearshore land slope, distance to lake center) metrics.} \label{table2}
  \centering
  % \setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{llll}
  \hline
  slope & distance & RMSE & $R^2$\\
  \hline
  true & true & - & -\\
  true & proxy & 4.2 m & 0.73\\
  proxy & true & 6.9 m & 0.26\\
  proxy & proxy & 6.6 m & 0.31\\
  \hline
\end{tabular}
\end{table}

@@

The most important covariates in offset models were those relating to spatial location, lake area, and perimeter (Figure S6). Conversely, watershed metrics and lake elevation had little contribution to random forest model fit (Figure S6). The spatial location (i.e. HUC4) covariate was notably less importance in the true slope model compared to the two proxy slope models. Model importance calculations indicated that omitting a geometric max depth measure results in a 130\%, 60\%, or 50\% increase in mean square error depending on the formulation of geometric max depth in Eq \ref{eq1} (Figure S6).

\section{Discussion}
\noindent
Our tests of geometric lake depth models show that specific proxy measures of lake geometry are not representative of true measures of lake geometry across a broad array of lakes. Using a cone model example, we show that nearshore land slope is not representative of in-lake slope. Furthermore, our results indicate that the use of nearshore land slope for prediction results in increased error and systematic overestimation of depth in concave and reservoir lakes. Although our analysis was limited to lakes with available bathymetry data, these lakes did not have characteristics that differed from that of the overall lake population (Figure S7, S2). This lack of difference suggests that our results are likely to be broadly applicable to all lakes although there is a possibility that there is some hidden bias not explored for in our analyses.

\subsection{Representativeness of proxy measures of lake geometry}
\noindent
In comparing among lake geometry measures, our analysis suggests that proxy distance to lake center is representative of true distance to the deepest point of the lakes but that proxy nearshore land slope is not representative of true in-lake slope. A simple indication of this non representativeness is that proxy nearshore land slope was often (in > 74\% of cases) steeper than true in-lake slope. This finding is consistent with \citeA{heathcotePredictingBathymetricFeatures2015} whos results suggest that in-lake slopes are shallower compared to the surrounding land. The shallow nature of in-lake slopes is likley a function of erosion and sediment transport processes \cite{hakansonLakeBottomDynamics1981, johanssonNewApproachesModelling2007}.

>>>>

\begin{document}
%TC:ignore
%% ------------------------------------------------------------------------ %%
%  Title
%
% (A title should be specific, informative, and brief. Use
% abbreviations only if they are defined in the abstract. Titles that
% start with general keywords then specific terms are optimized in
% searches)
%
%% ------------------------------------------------------------------------ %%

@@

\begin{keypoints}
\item Geometric models to predict lake depth, which require in-lake slope, assume that nearshore land slope is a good proxy for in-lake slope.
\item Using data from thousands of lakes, we show that nearshore land slope is a poor proxy for in-lake slope and increases prediction error.
\item Prediction errors were systematic such that depth was overpredicted in concave and reservoir lakes.
\end{keypoints}

@@

% The Plain Language Summary should be written for a broad audience,
% including journalists and the science-interested public, that will not have
% a background in your field.
%
% A Plain Language Summary is required in GRL, JGR: Planets, JGR: Biogeosciences,
% JGR: Oceans, G-Cubed, Reviews of Geophysics, and JAMES.
% see http://sharingscience.agu.org/creating-plain-language-summary/)
%
%% ------------------------------------------------------------------------ %%
%TC:endignore
%% \begin{abstract} starts the second page
\begin{abstract}
Lake depth is a critical characteristic that influences many important ecological processes in lakes. Unfortunately, lake depth measurements are labor-intensive to gather and are only available for a small fraction of lakes globally. Therefore, scientists have tried to predict lake depth from characteristics that are easily obtained for all lakes such as lake surface area or the slope of the land surrounding a lake. One approach for predicting lake depth simulates lake basins using a geometric model where nearshore land slope is assumed to be a representative proxy for in-lake slope and the distance to the center of the lake is assumed to be a representative proxy for the distance to the deepest point of the lake. However, these assumptions have rarely been tested in a broad range of lakes. We used bathymetry data from approximately 5,000 lakes and reservoirs to test these assumptions and to examine whether differences in lake type or shape influences depth prediction error. We found that nearshore land slope was not representative of in-lake slope and using it for prediction increases error substantially relative to models using true in-lake slope. We also found that models using nearshore land slope as a proxy systematically overpredict lake depth in concave lakes (i.e. bowl-shaped; up to 18\% of lakes in the study population) and reservoir lakes (up to 30\% of lakes in the study population), suggesting caution in using geometric models for depth prediction in unsampled lakes.
\end{abstract}

@@

\section{Introduction}
\noindent
\noindent
Lake depth is an important factor controlling lake physics, chemistry, and biota. Deeper lakes generally have higher water clarity and less complete mixing compared to shallow lakes \cite{feeEffectsLakeSize1996, readSimulating2368Temperate2014}. These differences are reflected in variation among lakes in terms of biological productivity \cite{qinWaterDepthUnderpins2020} and rates of greenhouse gas production \cite{liSignificantContributionLake2020}. However, because measured depth data is only available for a small fraction ($\sim$25\%) of all lakes, our ability to understand and predict depth-dependent processes is limited. The importance of lake depth, coupled with its limited availability, has led to numerous attempts to predict depth using measures available for all lakes such as lake surface area or the nearshore slope of the land surrounding a lake \cite{heathcotePredictingBathymetricFeatures2015, oliver2016prediction, sobekPredictingDepthVolume2011}. Given the limited prediction accuracy of prior depth prediction efforts ($\pm$ 6-7 m), a major focus has been on  improving prediction accuracy using strategies such as employing more diverse covariates \cite{oliver2016prediction}, varying lake buffer sizes \cite{heathcotePredictingBathymetricFeatures2015}, or estimating hidden groupings (e.g. fitting different models for distinct size classes) among lakes \cite{sobekPredictingDepthVolume2011}.

One approach for predicting lake depth involves using a geometric model that assumes lake basins correspond to an idealized shape such as a cone, bowl, or an elliptic sinusoid \cite{hollisterPredictingMaximumLake2011, neumannMaximumDepthAverage1959, yigzawNewGlobalStorage2018}. All such geometric models for lake depth prediction involve implicit assumptions about the terms of geometric formulae. In the simplest case, where lakes basins are treated as cones (Equation \ref{eq1}, Figure \ref{fig1}), two assumptions are required to make depth predictions for all lakes: 1) that nearshore land slope is a representative proxy for in-lake slope and 2) that the distance to the center of the lake is a representative proxy for the distance to the deepest point of the lake (Figure \ref{fig1}). This cone model imposes the following fixed (i.e. geometric) relationship between slope and horizontal distance:

@@

The assumptions of the cone model (as well as other geometric models) can be tested by comparing proxy measures of lake geometry against corresponding “true” (i.e. in-lake) values derived from bathymetric maps and by evaluating how lake cross-section shapes differ from that of an idealized cone \cite{johanssonNewApproachesModelling2007}. For instance, lake cross-section shapes have been shown to vary from narrow "convex" forms to outstretched "concave" forms \cite{hakansonLakeFormLake1977}. Because tests of geometric model assumptions require bathymetric map data, which is only available for about 15\% of all lakes, existing evidence may not be applicable to all lakes. The few studies that have tested these assumptions have been limited to individual studies of very large ($>$ 500 ha) lakes or studies on small numbers ($<$ 100) of lakes \cite{johanssonNewApproachesModelling2007}. Studies focused specifically on reservoirs (as opposed to the more typical case where reservoirs and natural lakes are combined), have been even more restricted to that of extremely large lakes $>$ 1000 ha \cite{lehnerHighresolutionMappingWorld2011, messagerEstimatingVolumeAge2016}.

As a result of this limited testing, we lack knowledge on both the predictive performance of geometric models, the effect of proxies on depth prediction, and whether depth predictions are more sensitive to measurement errors in the horizontal dimension (i.e. distance to the deepest point of the lake) or measurement errors in the vertical dimension (i.e. in-lake slope). Additionally it is unclear whether model prediction error is related to differences in lake type such those with different cross-section shapes or those classified as reservoirs versus natural lakes. Given these knowledge gaps, we asked three research questions: 1) How representative is nearshore land slope of in-lake slope; and how representative is the distance to the center of a lake compared to the distance to the deepest point of a lake? 2) How does the use of proxies for lake geometry affect lake depth prediction error? 3) How does lake cross-section shape (i.e. concave versus convex) and lake type (i.e. natural lake vs reservoir) affect depth prediction error? To answer these questions, we extracted maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, cross-section shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 lakes from bathymetric map data and supplemented this data with classification estimates of whether lakes are reservoirs or natural lakes. We used this data to compute geometric depth estimates (Equation \ref{eq1}) and prediction "offsets" to these estimates using the random forest algorithm (Equation \ref{eq3}). Covariates used in offset modeling included a variety of lake, watershed, and hydrologic subbasin measures that are available for all lakes (Table S1).

By definition, the distance proxy (distance to the center of the lake) must always be greater or equal to the true distance value (distance to the deepest point of the lake). Therefore, we expect that the use of this proxy will lead to overestimation of lake depth (Figure \ref{fig1}). Furthermore we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \cite{lanza1985interactions}. In a similar fashion, we expect to see overestimation error associated with using a nearshore land slope proxy in lakes with differing cross-section shape such that the depth of U-shaped (i.e. concave) lakes will be overpredicted whereas the depth of V-shaped (i.e. convex) lakes will be underpredicted (Figure S1). Finally, we expect that depth predictions themselves will be strongly related to lake area and hydrologic subbasin variables as these measures have been influential in prior studies \cite{oliver2016prediction}.

By testing these expectations, we can establish whether barriers to increased depth prediction accuracy lie in lack of correspondence between true and proxy measures of lake geometry or in hidden groupings among lakes (such as lake cross-section shape or reservoir status). This information could help direct future research efforts to focus on particular dimensions of lake geometry (i.e. horizontal versus vertical) or to stratify model predictions based on specific lake types and cross-section shapes. Ultimately, achieving increased depth prediction accuracy would allow for more precise estimates of depth-dependent biotic and chemical processes across broad spatial extents.

\section{Methods}
\subsection{Data description}
\noindent
We compiled bathymetry data on approximately 5,000 lakes in the Northeastern and Midwestern US from nine official state databases (Figure S2). The original data came in a variety of formats including pre-interpolated rasters (Minnesota), contour lines (Nebraska, Michigan, Massachusetts, Kansas, Iowa), contour polygons (New Hampshire, Connecticut), or point depth soundings (Maine). For the Minnesota data, we simply clipped the raster for each lake to its outline. For data from the remaining states, we processed each lake by converting its original representation to a point layer (if necessary), rasterizing these points, and creating an interpolated bathymetry “surface” using a simple moving window average in the \texttt{raster} R package \cite{hijmansRasterGeographicData2019}. The size of the moving window was adjusted iteratively to ensure that each bathymetry raster contained no missing data.

@@

We categorized lakes based on their cross-section shape and reservoir class. For cross-section shape, we categorized lakes as either convex or concave following the method of \citeA{hakansonLakeFormLake1977} by computing normalized lake depth-area relationships (i.e. hypsographic curves) and assigning class membership based on whether a lake’s curve falls above or below that of a simple straight-sided cone (Figure S3). We further classified lakes using data from \citeA{polus2020}, which uses the output of a machine learning algorithm to assign a probability to each lake as to whether it is a reservoir or a natural lake. For our purposes, we determined a lake to be a reservoir if the classification probability was 0.75 or greater. The \citeA{polus2020} data product defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water, beyond simply controlling water level. It makes no distinction between different dam types or dam heights.

Covariates used in random forest modeling (Table S1, Equation \ref{eq3}) for lake elevation, area, island area, perimeter, shoreline development, watershed to lake area ratio, and hydrologic subbasin (i.e. HUC4s), were obtained from the LAGOS-US LOCUS database \cite{smith2020}. One such measure, that of shoreline development, is a measure of lake perimeter shape defined as:

@@

\noindent
where sinuous lakes have larger values of shoreline development and circular lakes have smaller values of shoreline development. Watershed to lake area ratio is an approximation of water residence time and is defined as watershed area divided by lake area.

\subsubsection{Proxy evaluation}
\noindent
We conducted a qualitative assessment of whether or not proxy measures of lake geometry are representative of their true values by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination ($R^2$). We further tested proxy measures by examining their effect on lake depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of lake depth using only geometry information ($depth_{geometric}$, Equation \ref{eq1}). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all lakes (Table S1). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of $depth_{geometric}$ and among different lake types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table \ref{table2}).

@@

\subsubsection{Random forest models}
\noindent
Prior studies using geometric models to predict lake depth include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy. For our purposes, this offset modeling enabled us to test our expectations that prediction error would be different among different formulations of $depth_{geometric}$ and among different lake types. It also facilitated direct comparison against prior models of lake depth including those that are non-geometric. We generated an “offset” to geometric depth \cite<sensu>{hollisterPredictingMaximumLake2011} using the random forest algorithm and the \texttt{ranger} R package \cite{wrightRangerFastImplementation2017} to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation \ref{eq1}) along with the lake elevation, area, perimeter, and ratio/index measures listed in Table 1:

@@

\subsubsection{Model evaluations}
\noindent
We evaluated model fit and prediction error using root-mean-square error (RMSE) and coefficient of determination ($R^2$) metrics on a holdout set containing 25\% of all lakes. We evaluated the residuals of each model relative to lake cross-section shape and reservoir classes to determine whether depth is consistently over or under predicted for some lake types relative to others.

\section{Results}
\noindent
Lakes belonging to each cross-section shape and reservoir class were not evenly distributed across our study area (Figure S2, S3). For example, concave lakes were nearly absent from Michigan whereas Maine lakes had an overabundance of lakes categorized as neither concave nor convex. Lakes in the southern portions of our study area tended to be classified as reservoirs whereas lakes in the northern portions of our study area were a more even mix between reservoirs and natural lakes (Figure S2). Approximately 18\%, 80\%, and 2\% of lakes were classified as having a concave, convex, or neither shape respectively whereas approximately 30\% and 70\% of lakes were classified as being a reservoir or a natural lake.

Although proxy distance to lake center was often much larger in magnitude compared to the true distance to the deepest point of lakes’, they were strongly related ($R^2$ = 0.8). In contrast, proxy nearshore land slope and true in-lake slope were more weakly related ($R^2$ = 0.17). For slope measures, most lakes had lower magnitude (i.e. shallower) nearshore land slope compared to true in-lake slope (Figure \ref{fig3}). Taken together, these results suggest that proxy distance to the center of lakes is representative of true distance to the deepest point of lakes whereas proxy nearshore land slope is not representative of true in-lake slope. In addition to overall differences between slope and distance measures, we found differences in these relationships among lake shape classes. For example, in-lake slope and distance to the deepest point of the lake metrics were consistently larger in magnitude for convex lakes as compared to concave lakes (Figure S4). However,  there were not similar differences among slope and distance metrics for natural lakes versus reservoirs (Figure S4).

@@

\begin{table}[h]
  \caption{Model fit and predictive accuracy metrics (RMSE = root mean square error, $R^2$ = coefficient of determination) for all combinations of true (in-lake slope, distance to the deepest point of the lake) and proxy (nearshore land slope, distance to lake center) metrics.} \label{table2}
  \centering
  % \setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{llll}
  \hline
  slope & distance & RMSE & $R^2$\\
  \hline
  true & true & - & -\\
  true & proxy & 4.2 m & 0.73\\
  proxy & true & 6.9 m & 0.26\\
  proxy & proxy & 6.6 m & 0.31\\
  \hline
\end{tabular}
\end{table}

@@

The most important covariates in offset models were those relating to spatial location, lake area, and perimeter (Figure S5). Conversely, watershed metrics and lake elevation had little contribution to random forest model fit (Figure S5). The spatial location (i.e. HUC4) covariate was notably less importance in the true slope model compared to the two proxy slope models. Model importance calculations indicated that omitting a geometric max depth measure results in a 130\%, 60\%, or 50\% increase in mean square error depending on the formulation of geometric max depth in Equation \ref{eq1} (Figure S5).

\section{Discussion}
\noindent
Our tests of geometric lake depth models show that specific proxy measures of lake geometry are not representative of true geometry measures across a broad array of lakes. Models using non-representative proxies showed increased error and systematic overestimation of depth in concave and reservoir lakes. Although our analysis was limited to lakes with available bathymetry data, these lakes did not have characteristics that differed from that of the overall lake population (Figure S6, S8). Although there is a possibility that there is some hidden bias not explored for in our analyses, this lack of difference suggests that our results are likely to be broadly applicable to all lakes.

\subsection{Representativeness of proxy measures of lake geometry}
\noindent
In comparing among lake geometry measures, our analysis suggests that proxy distance to lake center is representative of true distance to the deepest point of the lakes but that proxy nearshore land slope is not representative of true in-lake slope. A simple indication of this non representativeness is that proxy nearshore land slope was often (in $>$ 74\% of cases) steeper than true in-lake slope. This finding is consistent with \citeA{heathcotePredictingBathymetricFeatures2015} whos results suggest that in-lake slopes are shallower compared to the surrounding land. The shallow nature of in-lake slopes is likley a function of erosion and sediment transport processes \cite{hakansonLakeBottomDynamics1981, johanssonNewApproachesModelling2007}.

One surprising finding with respect to the relationship between true and proxy geometry measures when examined by lake class was the fact that there was no greater difference between proxy and true distances in reservoirs compared to natural lakes. This is contrary to the idea that most reservoirs are drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \cite{lanza1985interactions}. One possible explanation is that the reservoir classification data from \citeA{polus2020} uses a more general definition of a reservoir (i.e. any permanent waterbody that has a water control structure likely to significantly impact flow or pool water) compared to that of conventional classifications that are tied to specific dam types or dam heights. Another possible explanation is that conventional reservoir classifications are conceptually biased towards more southern areas with few natural lakes (Figure S2).
